---
title: "Final_version_report"
author: "Guanghao He"
date: "3/12/2022"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
# import two data set and load all useful packages 
data_1 <- read.csv("~/Desktop/207/WHO-COVID-19-global-data.csv")
data_2 <- read.csv("~/Desktop/207/vaccination-data.csv")
library(dplyr)
library(lme4)
require(car)
require(MASS)
library(car)
library(ggplot2)
library(fitdistrplus)
library(logspline)
library(maps)
library(scales)
library(skimr)
library(snpar)
library(DHARMa)
library(glmmTMB)
library(boot)
library(htmlTable)
library(magrittr)
library(lawstat)
```



# I. Abstract

With the prevalence of COVID-19 and the arriving of vaccine, COVID-19 seems disappear gradually. However, we care more about the relationship between the cumulative death case, the number of vaccines received for different WHO regions. If there is no big area difference, the relationship between the number of vaccine types and the cumulative death case. We try to base on our data and model to answer our interest. Based on our model, we find that there isn't big difference between different WHO regions. Also, we can conclude that even with the increment of vaccine recieved, the cumulative death case for a country still increase, although our model fit well. This report consists of 6 main
parts: Introduction,Background, Method, Model Fit, Model diagnostics/ sensitivity analysis and Conclusion.

# II. Introduction

We are going to explore the influence towards the number of cumulative confirmed deaths reported to date (02/17/2022) of six levels of the factor different geographical regions of WHO regional offices(AFRO, AMRO, EMRO, EURO, SEARO, WPRO) and 10 levels of another factor types of used vaccine. Or to say specifically, why there exist the abnormal situation that more types of vaccines used, more cumulative deaths in different regions. As some regions and certain types of vaccine exists few sample and abnormal measurements, our final chosen factors are geographical regions with 4 levels ("AFRO",  "AMRO, "EURO", "WPRO") and the number of vaccine types used with 7 levels ("1", "2"..."7"). We use two data sets from [WHO Database](https://covid19.who.int/info). The first data file named daily cases and deaths by date reported to WHO and another is called Vaccination data. Our main data sets are published on the WHO website. The benefit of this public resource is that most data are clean and ready to be used. The shortcoming is that the WHO website doesn't provide more detailed information about the specific case. We can only learn the whole trend for different countries and different areas. Thus, we realize that the difference between areas plays an important role in spreading Covid and also due to the importance of vaccination to confirmed death cases. With the arrival of vaccination, more people get immunity to Covid. We decide to use two datasets: who-covid-19-global-data and vaccination data. Based on those two data sets, we can learn the updated cumulative confirmed case for different areas, vaccination receiving information. Because of the setting of our problem, we care more about the relationship between the cumulative death case, the number of vaccines received for different WHO regions. In our experiment, we assume there is no obvious difference between vaccines due to the lack of more specific information.

# III. Background

Although with the appearance of vaccines, our world has the weapon to restrict also confront with the Covid-19 virus, the reported infected cases and confirmed deaths still varies in different countries. The reason is multiple and we mainly focus on two factors. First is the number of used vaccine types of countries and second is different countries’ geographical regions based on WHO regional offices. In Figure () in the EDA part, we find that with the increment of vaccine used, the number of cumulative deaths still increases. This is more abnormal than our common sense. In common sense, the vaccine can help to build up resistance for the public and stop the spread of the Covid-19. That's why our interest comes into our mind.

# IV. Method
## 1. Data cleaning and processing 

The first step is to filter our data to the newest date because the number of confirmed cases and the number of vaccines received are cumulative. We only need to keep track of the newest update. The second step is to merge two datasets into one based on country. For each country, we will have the number of vaccines received, the cumulative confirmed cases, and belonging WHO regions. Finally, we remove all missing values because some countries lack death data. The reason for missing value is that some countries don't have too many citizens, which means these countries should be considered as extreme values. To make sure the accuracy of our future model, we simply remove those outliers. There is only three missing value for the number of vaccines types used, which doesn't influence the big picture. 

Important variables description are shown in below chart.

| column names               | explanation |
| :-----                     | :----- |
| WHO_region                 | The WHO divided the world into six region|
| NUMBER_VACCINES_TYPES_USED | The number of vaccines used |
| Cumulative_deaths          | The number of cumulative COVID-19 deaths |

## 2. Exploratory data analysis.

Before we go into details about our model, here is to show the severity trend of COVID-19 for the whole world. Below chart is the daily new COVID-19 case number and daily new COVID-19 death case aggregated all countries. The left y axis is the range of new daily case number, while the right y axis is the range of new daily death case number. The x axis is the same, which represented the date. From this plot, we find that the peak daily new case number happened at the same time of the peak daily death case number. Combined with below case mortality rate plot, between 2020/01 to 2020/03, COVID-19 cause more case mortality rate, which means if people infected by COVID-19 at that time, they face more risk to the death than people were infected recently. This can be explained by the prevalence of omicron. 

```{r,echo=FALSE,message=FALSE,warning=FALSE,fig.cap="Figure 1. The time series plot for daily new COVID-19 case number and daily new COVID-19 death case"}
daily_case <- data_1 %>% 
  group_by(Date_reported) %>% 
  mutate(daily_case = sum(New_cases),
         daily_death = sum(New_deaths)) %>%
  mutate(case_mortaility = daily_death*100/daily_case) %>%
  distinct(Date_reported,daily_case,daily_death,case_mortaility) 
scale_value = max(daily_case$daily_case, na.rm = TRUE)/ max(daily_case$daily_death, na.rm = TRUE)
ggplot(daily_case) %>% + 
  geom_line(aes(x=as.Date(Date_reported), y=daily_case), color = 'dodgerblue') +
  geom_line(aes(x=as.Date(Date_reported), y=daily_death*scale_value),color ='red') + 
  scale_y_continuous(sec.axis = sec_axis(~./scale_value, name = 'Daily death cases')) +
  xlab('Date reported') +
  ylab('Daily case number') +
  theme(axis.line.y.right = element_line(color = "red"), 
        axis.ticks.y.right = element_line(color = "red"),
        axis.text.y.right = element_text(color = "red"), 
        axis.title.y.right = element_text(color = "red"),
        axis.line.y.left = element_line(color = "dodgerblue"),
        axis.ticks.y.left  = element_line(color = "dodgerblue"),
        axis.text.y.left = element_text(color = "dodgerblue"), 
        axis.title.y.left = element_text(color = "dodgerblue")
        )
```

```{r,echo=FALSE,message=FALSE,warning=FALSE,fig.cap="Figure 2. The case mortality rate over time aggregated all countries"}
ggplot(daily_case)+ 
  geom_line(aes(x=as.Date(Date_reported), y=case_mortaility), color = 'dodgerblue') +
  xlab('Date reported') +
  ylab('Daily case mortality rate over time') 

```


The first step for our analysis is to summary all related variables. We combine two data set based on the country. Below is the result of summary statistics of data set merged by original two data sets. There are 223 observation in total with 19 columns. This is not our final data sets because of the existence of missing values. From below summary, we could find the basic statistics summary of each column like mean, maximum, standard deviation and etc. Also, the histogram of each column is attached in the following table. To be notice, to answer our interest, we should consider removing the missing values and 0. At this part, we don't specific this process.

```{r,echo=F,message=FALSE,warning=FALSE}
data_1 <- data_1 %>% filter(data_1$Date_reported == '2022-02-17')
colnames(data_2)[colnames(data_2)=="COUNTRY"] <- "Country"
data_3 <- merge(data_1,data_2,by.x="Country") %>%  dplyr::select(-c(ISO3,WHO_REGION))
# colnames(data_3)[colnames(data_3)=="锘緿ate_reported"] <- "Date_reported"
skim(data_3)
```

The next step for EDA is to show the world trennd of COVID-19 based on each country's cumulative death cases. To be notice, there are some missing countries in this plot because of the lack of longitude and latitude information of those countries. Thus, we cannot show those countries like the United States. However, this plot still help us to show the overall trend of cumulative case in the world.

```{r,echo=FALSE,message=FALSE,warning=FALSE,fig.cap="Figure 3. The map of World's cumulative death cases."}
world <- map_data("world")
worldplot <- ggplot() +
  geom_polygon(data = world, aes(x=long, y = lat, group = group)) 

covid.today<- data_3 %>% 
  filter(Date_reported == "2022-02-17") %>% 
  mutate(region=Country)

covid.today.world<- inner_join(world, covid.today, by = "region")

ggplot() +
  geom_polygon(data = covid.today.world) +
  aes(x=long, y = lat, group = group,fill=Cumulative_cases) + 
  coord_fixed(1.3) +scale_fill_viridis_c(option = "D", labels=comma)
```

Third step is to filter our target data sets that satisfy our needs. So, we only keep three columns: WHO_region, NUMBER_VACCINES_TYPES_USED and Cumulative_deaths. The detail explanation for each variables is shown in previous section. The missing value and abnormal measurement for number of vaccine type used is 3. The specific reason why we remove those missing values are explained in data cleaning parts. 

```{r,echo=F}
data <- data_3 %>% dplyr::select(c(WHO_region,NUMBER_VACCINES_TYPES_USED,Cumulative_deaths))
data[data==0] <- NA
skim(data)
data$WHO_region <- as.factor(data$WHO_region)
data$NUMBER_VACCINES_TYPES_USED <- as.factor(data$NUMBER_VACCINES_TYPES_USED)
```

Here we shows the frequency table to show the reason why we shouldn't to use Two-way ANOVA model as our model candidate. There exists 0 observation in certain level of two factors, which is discrete or like counts and hence two-way ANOVA may not be a good choice. The generalized linear mixed models is much more fitful model in theoretical analysis.

Generate frequency tables:
```{r,echo=F}
table(data$WHO_region,data$NUMBER_VACCINES_TYPES_USED)
```

The left frequency table is shown below. We observe there doesn't 0 in our data set, which stratify our need for model selection and the quality of choose levels of two factors are better. 

```{r,echo=F}
target <- c("AFRO","AMRO","EURO","WPRO")
data_chs <-  data %>% filter(WHO_region%in% target)
# table(as.character(data_chs$WHO_region),data_chs$NUMBER_VACCINES_TYPES_USED)
target <- c("1","2","3","4","5","6","7")
data_chs <- data_chs %>% filter(NUMBER_VACCINES_TYPES_USED %in% target)
table(as.character(data_chs$WHO_region),as.character(data_chs$NUMBER_VACCINES_TYPES_USED))
data_chs <- na.omit(data_chs)
```

Now, we explore our response variable's distribution. We use the Box-Cox transformation to apply remedy way of asymmetry of a variable, different variances or non linearity between variables. When there appears to be unequal variance or/and non-normality problems, transformation should be done on the response variable Y. Below is the result of Box-cox transformation. From this figure, we find that the 0 is inside the 95% confidence interval of the optimal λ and as the estimation of the parameter is really close to 0, the best option is to apply the logarithmic transformation of the data. 

```{r,message=FALSE,warning=FALSE,echo=FALSE,fig.cap="Figure 4. The Boxcox plot of response variable"}
boxcox(data_chs$Cumulative_deaths~data_chs$WHO_region+data_chs$NUMBER_VACCINES_TYPES_USED)
```

Then, we plot for the box plot of the number of vaccine types used vs the log(Cumulative death case). As we talk in the Background part, the abnormal incresment come into our mind and violate common sense. Following EDA will show how we exclude other models candidate and finally find our final model.   

```{r,message=FALSE,warning=FALSE,echo=FALSE,fig.cap="Figure 5. The boxplot of number of vaccine used vs the log(Cumulative death case)"}
ggplot(data_chs %>% filter(!is.na(NUMBER_VACCINES_TYPES_USED)), aes(x=factor(NUMBER_VACCINES_TYPES_USED),fill = as.factor(NUMBER_VACCINES_TYPES_USED), y = log( Cumulative_deaths))) + 
  geom_boxplot()+
  labs(fill = "Vaccine types used",
       x= "Vaccine types used",
       y= "Cumulative deaths",
       title= "The boxplot of vaccine types and cumulative deaths")
```

```{r,message=FALSE,warning=FALSE,echo=FALSE,fig.cap= "Figure 6.The histogram of number of vaccine used vs the log(Cumulative death case)"}
hist(log(data_chs$Cumulative_deaths),main="Histogram of cumulative death", xlab="cumulative death")
```

Below the pairwise scatter plot, we could see there is no obvious relationship between our three variables. 

```{r,message=FALSE,warning=FALSE,echo=FALSE,fig.cap="Figure 7. The pairwise scatter plot among three variables."}
pairs(data_chs%>% ungroup() %>%dplyr::select(NUMBER_VACCINES_TYPES_USED,WHO_region,Cumulative_deaths),pch=16,col='red',cex=1.5)
```

Before fitting the model, we would like to investigate whether there is linear relationship between depedent variables and indepedent variables by drawing the scatter plot. As we mentioned before, it is important to take log transformation of origianl variables.

Below is the scatter plot of the number of vaccine used and cumulative deaths. 
Most region have the trend is almost the same, pointing to top-right. To be notice, the number of cumulative death is log transformation. At this point, we have a sense to use multi-level models (or GLM)

```{r,message=FALSE,warning=FALSE,echo=FALSE,fig.cap="Figure 8. The Scatter plot of number of vaccine used vs the log(Cumulative death case)"}
ggplot(data_chs) + 
  aes(y = log(Cumulative_deaths), 
      x = NUMBER_VACCINES_TYPES_USED, 
      colour = WHO_region,
      group = as.factor(WHO_region)) +
  geom_point() + 
  geom_smooth(aes(color = as.factor(WHO_region)),method = "lm", se=FALSE, fullrange=TRUE) +
  labs(title = "Number of vaccine types used vs log(Cumulative deaths)",
       x="Number of vaccine types used",
       y="log(Cumulative deaths)",
       fill= 'WHO region')
```


## 3. Model Selection

Then, we want to estimate the necessary parameters to describe the assumed error distribution. The following part is used for selecting error distribution candidate. The Cullen and Frey Graph below is a good way to exempt some distributions by the parameters of skewness and kurtosis using the descdist function. We find that our observation (blue point) seems belong to normal,beta and Weibull distribution. It's worth noting that all cumulative deaths cases is log transformation. 

```{r,echo=F,fig.cap="Figure 8. The Cullen and Frey graph for log(Cumulative death case)"}
descdist(log(data_chs$Cumulative_deaths), discrete = F)
```

```{r,echo=FALSE}
fit.norm <- fitdist(log(data_chs$Cumulative_deaths), "norm")
range01 <- function(x){(x-min(x)*1.01)/(max(x)*1.01-min(x)*1.01)}
fit.beta <- fitdist(abs(range01(log(data_chs$Cumulative_deaths))), "beta")
fit.weibull <- fitdist(log(data_chs$Cumulative_deaths), "weibull")
```

The next step is to separatlly fit for each type of distribution to investigate the true distribution. We seperatly fit for beta, normal and Weibull distribution. The specific checking graphs are shown in appendix. We find the QQ plot for beta distribution fit better than other two distribution. We based on the minimum AIC rule to select beta distribution and normal distribution as candidates. To be notice,we performed the normalization to make sure the beta value between 0 and 1. Below is graph for AIC value.

```{r,echo=FALSE}
matrix(c(fit.beta$aic,fit.weibull$aic,fit.norm$aic),
       ncol = 1,
       dimnames = list(c("beta distribution", "weibull distribution","normal distribution"),c("AIC value"))) %>% 
  htmlTable
```

# V. Model Fit
## 1. Final selection

Current stage, we have three candidates: linear mixed-effect model with normal distribution response, generalized linear model with beta distribution response and generalized linear mixed model with beta distribution response. 

The final model that we choose is the generalized linear model with beta distribution response. The final model is shown below. 

```{r}
GLM_2 <- glmmTMB(abs(range01(log(Cumulative_deaths)))~ NUMBER_VACCINES_TYPES_USED,family=beta_family(link="logit"), data=data_chs)
```

And to see the below chart, all variables are significant at alpha = 0.05 level and the value is round to two digits.
```{r,echo=FALSE}
matrix(c(round(summary(GLM_2)[["coefficients"]][["cond"]][,1],2),round(summary(GLM_2)[["coefficients"]][["cond"]][,4],3)),
       ncol = 2,
       dimnames = list(c("Intercept","NUMBER_VACCINES_TYPES_USED2","NUMBER_VACCINES_TYPES_USED3","NUMBER_VACCINES_TYPES_USED4","NUMBER_VACCINES_TYPES_USED5","NUMBER_VACCINES_TYPES_USED6","NUMBER_VACCINES_TYPES_USED7"),
                       c("Estimate","Pr(>|z|)")))%>% 
  htmlTable
```

The reason why we choose this model is from two aspects. I attached other candidate formula in appendix part. The first aspect is to the failure of random effect. From previous EDA figure, we find that it seems like different WHO regions have same slope but differnt intercept, pointing to top-right. We would like to choose WHO region as random effect and the number of vaccine received as fixed effect. However, we could figure out the random effect of WHO region is not highly significant, as the random effect could only count for $0.69/(0.69+5.23)=11%$ of the total variance, which means this effect is not meaningful. From the fix effect estimates, we could see the the first type and the seven types of vaccine used are more statistically significant in outcome variable. Thus, we decide to remove our random effect and also remove the WHO region as our variable candidates. 

The second aspect is to use beta link function. Since we already remove WHO region, our previous linear mixed model become to simple linear model. Our generalized linear mixed model become to generalized linear model. We compared with AIC and BIC value for normal distribution of simple linear regression and the beta distribution of generalized linear model. Below is the AIC and BIC value for two candidate. We find generalized linear model both have smaller AIC and BIC value than another one. Thus, our final model is generalized linear model with beta link function. 

```{r,echo=FALSE}
# linear mixed model 
lmm <- lmer(log(Cumulative_deaths)~ NUMBER_VACCINES_TYPES_USED + (1 | WHO_region), data = data_chs, REML = T)
lmm_2 <- lm(log(Cumulative_deaths)~ NUMBER_VACCINES_TYPES_USED, data = data_chs)
# generalized linear mixed model with beta distribution response
GLM <- glmmTMB(abs(range01(log(Cumulative_deaths)))~ NUMBER_VACCINES_TYPES_USED + (1 | WHO_region),family=beta_family(link="logit"), data=data_chs)
```


```{r,echo=FALSE}
matrix(c(round(AIC(lmm_2),2),round(AIC(GLM_2),2),round(BIC(lmm_2),2),round(BIC(GLM_2),2)),
       ncol = 2,
       dimnames = list(c("Simple linear model","generalized linear model"),
                       c("AIC","BIC")))%>% 
  htmlTable
```

## 2. Model Notation

$$
log(Y) = -1.03 + 0.69 X_2 + 1.04 X_3 +1.35 X_4 +1.37 X_5 +1.24 X_6 +1.73 X_7
$$
Y is normailzed from 0 to 1. Based on the slope coefficient of our model, we set the country which recived 1 dose of vaccine as the reference group. Compared with country which recived 1 dose of vaccine, the death number of country which recived 2 dose of vaccine increase 99.4%. Compared with country which recived 1 dose of vaccine, the death number of country which recived 3 dose of vaccine increase 182.9%. Compared with country which recived 1 dose of vaccine, the death number of country which recived 4 dose of vaccine increase 285.7%. Compared with country which recived 1 dose of vaccine, the death number of country which recived 5 dose of vaccine increase 294.5%.Compared with country which recived 1 dose of vaccine, the death number of country which recived 6 dose of vaccine increase 245.6%. Compared with country which recived 1 dose of vaccine, the death number of country which recived 7 dose of vaccine increase 464.0%.

Thus, this model means that compared with countries with only 1 type of vaccine, other countries have different rate of death increasment. This can be explained opposite with our common senese, which although the types of vaccine increase, the death number still increase. Because those countries only get only 1 or 2 types of vaccince are developing countries which have less citizens than developed countries. The prevelance of COVID-19, unfortunatl, doesn't stop by the increase of vaccine type. It still meet with parts of our common sense. It can be explained by the degree of development of countries, developed countries have more medical resource and more citizens, which still lead to the large number of death because of large infected number.

Also, combined with our previous model selection process, we can conclude there is no big difference between WHO regions aimed for the cumulative death number and the number of vaccine type. 

## 3. Model assumption 

Our model has assumptions from below aspects: 1. The data are independently distributed 2. A GLM does NOT assume a linear relationship between the response variable and the explanatory variables, but it does assume a linear relationship between the transformed expected response in terms of the link function and the explanatory variables. 3.The homogeneity of variance does NOT need to be satisfied. 

Standard GLMs are agnostic to the causal structure of the data to which they are fitted. The process of fitting a GLM makes no assumptions about causality, nor does it enable any conclusions about causality to be drawn without further strong assumptions. 

Under our model, no assumptions are violated. From our finding distribution process, we find our response variable have beta distribution. After the log transformation, we find that there is a linear relationship between the transformed expected response and the independent variables. Moreover, our residual plots show the normal distribution of our residual. Thus, there is no specific assumption violated for our model. 

# VI. Model diagnostics and sensitivity analysis 

## 1. Deviance Residuals and Pearson Residuals

The boxplot shows similar distributions of the two types of residuals, so this particular
diagnostic does not provide any indication for lack-of-fit.
```{r,echo=FALSE,fig.cap="Figure 9 The residual distribution of residuals"}
res.P = residuals(GLM_2, type="pearson")
res.D = residuals(GLM_2)
boxplot(cbind(res.P, res.D), names = c("Pearson", "Deviance"))
```

## 2. Residual Plots

The quantile plot (also called a normal probability plot) does not raise any significant concern with normality of the weighted residuals.

```{r,echo=FALSE,fig.cap="Figure 10 normal qq plot for final model"}
qqnorm(resid(GLM_2))
qqline(resid(GLM_2)) 
```

The plot of residuals against ftted values seems randomly distributed around 0, which does not raise any significant concern to our goodness- fit. 
```{r,echo=FALSE,fig.cap="Figure 11 residuals against fitted values"}
plot(fitted(GLM_2),resid(GLM_2))
abline(h=0, col="blue")
```

## 3. Runs Test

Below is the result of runs test. The null hypothesis of runs test that there are no systematic patterns in a sequence of random numbers, here the residuals.Any systematic trend (no matter positive/negative correlated) suffices to reject the goodness-of-fit of the model. Here our p-value is large enough which fail to reject the null hypothesis. This means our model has goodness-of-fit. 

```{r,echo=FALSE}
runs.test(y = res.D, plot.it = TRUE)
```

# VII. Conclusion

Based on our model and our analysis, unfortunately, the cumulative death for countries doesn't decrease with the increase of vaccine type. Also, since we exclude the area difference of WHO regions,this is opposite with our commmon sense. However, it still meet with parts of our common sense. It can be explained by the degree of development of countries, developed countries have more medical resource and more citizens, which still lead to the large number of death because of large infected number. Although our model fit the data well without violating any assumptions and without any causal inference, our hope that increasing the vaccine type can stop the COVID-19 is impossible. 

From the beginning EDA part, we find that it is impossible to answer our interest by two-way ANOVA. Then, we based on three model candidates based on two aspects to choose the best model for our data, which is the generalized linear model with beta distribution response. One is due to the failure of randome effect. Another one is due to the link function. Last step is to do model diagnostics and test for the assumption. Our model fit our data very without violating any assumption. 

In conclusion, we successfully choose a approriate model to answer our interest, even if our answer for our interest is hopeless. 

# Data Source

1.Data_1
https://covid19.who.int/WHO-COVID-19-global-data.csv

2.Data_2
https://covid19.who.int/who-data/vaccination-data.csv

# APPENDIX
## 1. Check for response variable's distribution
### a. normal distribution 
```{r,echo=FALSE}
plot(fit.norm)
```

### b. beta distribution 
```{r,echo=F}
plot(fit.beta)
```

### c. Weibull distribution

```{r,echo=F}
plot(fit.weibull)
```

## 2. Other model candidates
```{r}
# linear mixed model 
lmm <- lmer(log(Cumulative_deaths)~ NUMBER_VACCINES_TYPES_USED + (1 | WHO_region), data = data_chs, REML = T)
lmm_2 <- lm(log(Cumulative_deaths)~ NUMBER_VACCINES_TYPES_USED, data = data_chs)
# generalized linear mixed model with beta distribution response
GLM <- glmmTMB(abs(range01(log(Cumulative_deaths)))~ NUMBER_VACCINES_TYPES_USED + (1 | WHO_region),family=beta_family(link="logit"), data=data_chs)
```


# Code attached 
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```